---
title: "Impact of Enhanced Seasonal Malaria Chemoprevention on Hemoglobin Trajectories in Children: A Randomized Controlled Trial Analysis"
author: 
  - name: "Applied Data Analysis Consulting Group"
    affiliation: "African Institute for Mathematical Sciences — Rwanda"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 8
    fig-height: 5
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Libraries 
library(dplyr)
library(lme4)
library(lmerTest)
library(knitr)
library(kableExtra)
library(broom.mixed)
library(patchwork)
library(ggplot2)

# Load & prepare data
dat <- read.csv("malaria_anemia_rct.csv")

dat <- dat %>%
  mutate(
    treatment_f = factor(treatment, levels = c(0, 1),
                         labels = c("Standard SMC", "Enhanced SMC")),
    itn_use_f   = factor(itn_use, levels = c(0, 1),
                         labels = c("Inconsistent", "Consistent")),
    child_id    = factor(child_id)
  )


trt_colors <- c("Standard SMC" = "#E74C3C", "Enhanced SMC" = "#2ECC71")
```

---

# Exploratory Data Analysis

## Dataset Overview

```{r data-overview}
# Basic counts 
n_children <- length(unique(dat$child_id))
n_obs      <- nrow(dat)
n_miss     <- sum(is.na(dat))

tibble(
  Characteristic = c(
    "Total observations", "Number of children",
    "Time points (months)", "Missing values",
    "Standard SMC (n children)", "Enhanced SMC (n children)"
  ),
  Value = c(
    n_obs, n_children,
    paste(sort(unique(dat$month)), collapse = ", "),
    n_miss,
    length(unique(dat$child_id[dat$treatment == 0])),
    length(unique(dat$child_id[dat$treatment == 1]))
  )
) %>%
  kable(align = "lr", caption = "Dataset Structure Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Baseline Characteristics and Randomization Check

```{r baseline-table, warning=false}
#  Baseline summary by treatment arm 
baseline <- dat %>% filter(month == 0)

baseline %>%
  group_by(treatment_f) %>%
  summarise(
    N = n(),
    `Age — mean (SD)`        = paste0(round(mean(age_baseline), 1), " (",
                                      round(sd(age_baseline),   1), ")"),
    `Hemoglobin — mean (SD)` = paste0(round(mean(hemoglobin), 2), " (",
                                      round(sd(hemoglobin),   2), ")"),
    `Consistent ITN use (%)`  = paste0(round(mean(itn_use) * 100, 1), "%"),
    .groups = "drop"
  ) %>%
  kable(
    col.names = c("Treatment Arm", "N", "Age (months)",
                  "Hemoglobin (g/dL)", "ITN Use (%)"),
    align = "lcccc",
    caption = "Baseline Characteristics by Treatment Group"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r randomization-tests, warning=false}
#  Statistical tests for baseline balance 
t_age   <- t.test(age_baseline ~ treatment, data = baseline)
t_hb    <- t.test(hemoglobin  ~ treatment, data = baseline)
chi_itn <- chisq.test(table(baseline$treatment_f, baseline$itn_use_f))

tibble(
  Variable = c("Age at baseline", "Hemoglobin at baseline", "ITN use"),
  Test     = c("Two-sample t-test", "Two-sample t-test", "Chi-square test"),
  `p-value` = round(c(t_age$p.value, t_hb$p.value, chi_itn$p.value), 3)
) %>%
  kable(align = "llc", caption = "Randomization Balance Tests") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Hemoglobin Distributions Over Time

```{r boxplots}
#  Boxplots by time point and treatment arm 
ggplot(dat, aes(x = factor(month), y = hemoglobin, fill = treatment_f)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  scale_fill_manual(values = trt_colors, name = "Treatment") +
  labs(
    x = "Follow-up Month",
    y = "Hemoglobin (g/dL)",
    title = "Hemoglobin Distribution by Time Point and Treatment Arm"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

## Individual Trajectories (Spaghetti Plot)

```{r spaghetti-plot, fig.height=6, warning=FALSE}
#  Individual lines + LOESS mean overlay 
ggplot(dat, aes(x = month, y = hemoglobin,
                group = child_id, color = treatment_f)) +
  geom_line(alpha = 0.15, linewidth = 0.3) +
  geom_smooth(
    aes(group = treatment_f),
    method  = "loess",
    se      = FALSE,
    linewidth = 1.6,
    formula = y ~ x
  ) +
  scale_color_manual(values = trt_colors, name = "Treatment") +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  labs(
    x        = "Follow-up Month",
    y        = "Hemoglobin (g/dL)",
    title    = "Individual Hemoglobin Trajectories with LOESS Mean"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

## Mean Trajectories with 95% Confidence Bands

```{r mean-trajectories}
#  Observed group means ± 1.96 SE 
mean_dat <- dat %>%
  group_by(treatment_f, month) %>%
  summarise(
    mean_hb = mean(hemoglobin, na.rm = TRUE),
    se_hb   = sd(hemoglobin,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(mean_dat,
       aes(x = month, y = mean_hb, color = treatment_f, fill = treatment_f)) +
  geom_ribbon(
    aes(ymin = mean_hb - 1.96 * se_hb,
        ymax = mean_hb + 1.96 * se_hb),
    alpha = 0.15, color = NA
  ) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = trt_colors, name = "Treatment") +
  scale_fill_manual( values = trt_colors, name = "Treatment") +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  labs(
    x     = "Follow-up Month",
    y     = "Mean Hemoglobin (g/dL)",
    title = "Observed Mean Hemoglobin Trajectories with 95% Confidence Bands"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```



# Model Building and Selection

## Unconditional Means Model and ICC

```{r null-model ,warning = false}
# random intercept
m0 <- lmer(hemoglobin ~ 1 + (1 | child_id), data = dat, REML = TRUE)
summary(m0)
vc0          <- as.data.frame(VarCorr(m0))
var_child    <- vc0$vcov[1]
var_residual <- vc0$vcov[2]
icc_val      <- var_child / (var_child + var_residual)

tibble(
  Component              = c("Between-child variance (τ₀²)",
                             "Within-child variance (σ²)",
                             "Intraclass Correlation Coefficient (ICC)"),
  Value = round(c(var_child, var_residual, icc_val), 3)
) %>%
  kable(align = "lr", caption = "Unconditional Means Model — Variance Components") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Fixed Effects Model Comparison

```{r fixed-effects-models, warning=false}
#add fixed effects (Model comparison)
m1 <- lmer(hemoglobin ~ month + (1 | child_id),
           data = dat, REML = FALSE)

m2 <- lmer(hemoglobin ~ month + treatment_f + (1 | child_id),
           data = dat, REML = FALSE)

m3 <- lmer(hemoglobin ~ month + treatment_f + age_baseline + itn_use_f +
             (1 | child_id),
           data = dat, REML = FALSE)

m4 <- lmer(hemoglobin ~ month * treatment_f + age_baseline + itn_use_f +
             (1 | child_id),
           data = dat, REML = FALSE)
summary(m1)
summary(m2)
summary(m3)
summary(m4)
tibble(
  Model  = c("M1: Month only",
             "M2: + Treatment",
             "M3: + Age + ITN use",
             "M4: + Treatment × Month (interaction)"),
  df     = c(attr(logLik(m1), "df"), attr(logLik(m2), "df"),
             attr(logLik(m3), "df"), attr(logLik(m4), "df")),
  AIC    = round(c(AIC(m1), AIC(m2), AIC(m3), AIC(m4)), 1),
  BIC    = round(c(BIC(m1), BIC(m2), BIC(m3), BIC(m4)), 1),
  logLik = round(c(as.numeric(logLik(m1)), as.numeric(logLik(m2)),
                   as.numeric(logLik(m3)), as.numeric(logLik(m4))), 1)
) %>%
  kable(align = "lrrrr",
        caption = "Fixed Effects Model Comparison (ML estimation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(4, bold = TRUE, background = "#E8F8E8")
```

## Random Effects Structure Comparison

```{r random-effects-comparison , warning=false}
#  Random intercept only vs. random intercept + slope 
m_ri  <- lmer(hemoglobin ~ month * treatment_f + age_baseline + itn_use_f +
                (1 | child_id),
              data = dat, REML = FALSE)

m_ris <- lmer(hemoglobin ~ month * treatment_f + age_baseline + itn_use_f +
                (1 + month | child_id),
              data = dat, REML = FALSE)
summary(m_ri)
summary(m_ris)
#  AIC/BIC comparison table 
tibble(
  Model  = c("Random Intercept Only",
             "Random Intercept + Random Slope (month)"),
  df     = c(attr(logLik(m_ri),  "df"), attr(logLik(m_ris), "df")),
  AIC    = round(c(AIC(m_ri),    AIC(m_ris)), 1),
  BIC    = round(c(BIC(m_ri),    BIC(m_ris)), 1),
  logLik = round(c(as.numeric(logLik(m_ri)), as.numeric(logLik(m_ris))), 1)
) %>%
  kable(align = "lrrrr",
        caption = "Random Effects Structure Comparison (ML estimation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r lrt}
#  Likelihood Ratio Test 
lrt <- anova(m_ri, m_ris)
lrt
```

## Final Model (REML)

```{r final-model}
# Final model
m_final <- lmer(
  hemoglobin ~ month * treatment_f + age_baseline + itn_use_f +
    (1 + month | child_id),
  data = dat, REML = TRUE
)

summary(m_final)
```

---

# Model Diagnostics

```{r diagnostics, fig.height=7}
#   residuals and random effects
resid_vals  <- residuals(m_final)
fitted_vals <- fitted(m_final)
re          <- ranef(m_final)$child_id

#  Residuals vs Fitted 
p1 <- ggplot(data.frame(fitted = fitted_vals, resid = resid_vals),
             aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue",
              linewidth = 0.8, formula = y ~ x) +
  labs(x = "Fitted Values", y = "Residuals",
       title = "(A) Residuals vs. Fitted") +
  theme_minimal(base_size = 11)

# Q-Q — observation-level residuals 
p2 <- ggplot(data.frame(resid = resid_vals), aes(sample = resid)) +
  stat_qq(alpha = 0.3, size = 0.7) +
  stat_qq_line(color = "red") +
  labs(title = "(B) Q-Q Plot: Level-1 Residuals") +
  theme_minimal(base_size = 11)

# Q-Q — random intercepts 
p3 <- ggplot(data.frame(intercept = re[, 1]), aes(sample = intercept)) +
  stat_qq(alpha = 0.5) +
  stat_qq_line(color = "red") +
  labs(title = "(C) Q-Q Plot: Random Intercepts") +
  theme_minimal(base_size = 11)

#  Panel D: Q-Q — random slopes 
p4 <- ggplot(data.frame(slope = re[, 2]), aes(sample = slope)) +
  stat_qq(alpha = 0.5) +
  stat_qq_line(color = "red") +
  labs(title = "(D) Q-Q Plot: Random Slopes") +
  theme_minimal(base_size = 11)

(p1 + p2) / (p3 + p4)
```



# Confirmatory Analysis

#Fixed Effects Results Table

```{r results-table}
#  table from the REML final model
tidy_final <- tidy(
  m_final,
  effects = "fixed",
  conf.int = TRUE,
  conf.level = 0.95
)

tidy_final %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "month" ~ "Month",
      term == "treatment_fEnhanced SMC" ~ "Treatment (Enhanced vs. Standard)",
      term == "age_baseline" ~ "Age at Baseline (months)",
      term == "itn_use_fConsistent" ~ "ITN Use (Consistent vs. Inconsistent)",
      term == "month:treatment_fEnhanced SMC" ~ "Month × Treatment (Interaction)",
      TRUE ~ term
    )
  ) %>%
  select(
    Term = term,
    Estimate = estimate,
    SE = std.error,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `t-value` = statistic,
    `p-value` = p.value
  ) %>%
  mutate(
    across(Estimate:`t-value`, ~ round(., 3)),
    `p-value` = ifelse(`p-value` < 0.001, "< 0.001",
                       sprintf("%.4f", `p-value`))
  ) %>%
  kable(align = "lrrrrrr") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(6, bold = TRUE, background = "#FFF9C4")
```

#Variance Components

```{r variance-components , warning=false}
#  Random effects variance components 
vc_final <- as.data.frame(VarCorr(m_final))

vc_final %>%
  select(Group = grp, Effect = var1, Variance = vcov, SD = sdcor) %>%
  mutate(
    Group  = ifelse(Group == "child_id", "Child (Level 2)", "Residual (Level 1)"),
    Effect = case_when(
      Effect == "(Intercept)" ~ "Random Intercept",
      Effect == "month"       ~ "Random Slope (Month)",
      is.na(Effect)           ~ "Covariance",
      TRUE                    ~ Effect
    ),
    Variance = round(Variance, 4),
    SD       = round(SD,       4)
  ) %>%
  kable(
    col.names = c("Level", "Random Effect", "Variance", "Std. Dev."),
    align     = "llrr",
    caption   = "Variance Components of the Final Linear Mixed Model"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

# Model-Based Predicted Trajectories

```{r predicted-trajectories}
#Fixed-effects predictions at mean covariate values 
itn_ref <- names(sort(table(dat$itn_use_f), decreasing = TRUE))[1]

pred_grid <- expand.grid(
  month       = seq(0, 6, by = 0.5),
  treatment_f = factor(c("Standard SMC", "Enhanced SMC"),
                       levels = c("Standard SMC", "Enhanced SMC")),
  age_baseline = mean(dat$age_baseline, na.rm = TRUE),
  itn_use_f    = factor(itn_ref, levels = levels(dat$itn_use_f))
)

mm       <- model.matrix(~ month * treatment_f + age_baseline + itn_use_f,
                         data = pred_grid)
fe       <- fixef(m_final)
pred_grid$pred  <- as.numeric(mm %*% fe)
pred_se         <- sqrt(diag(mm %*% vcov(m_final) %*% t(mm)))
pred_grid$lower <- pred_grid$pred - 1.96 * pred_se
pred_grid$upper <- pred_grid$pred + 1.96 * pred_se

#  Observed means for reference points 
obs_means <- dat %>%
  group_by(treatment_f, month) %>%
  summarise(mean_hb = mean(hemoglobin, na.rm = TRUE), .groups = "drop")

ggplot(pred_grid,
       aes(x = month, y = pred, color = treatment_f, fill = treatment_f)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.3) +
  geom_point(data   = obs_means,
             aes(x  = month, y = mean_hb, color = treatment_f),
             size   = 3, shape = 16, inherit.aes = FALSE) +
  scale_color_manual(values = trt_colors, name = "Treatment") +
  scale_fill_manual( values = trt_colors, name = "Treatment") +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  labs(
    x        = "Follow-up Month",
    y        = "Predicted Hemoglobin (g/dL)",
    title    = "Model-Based Predicted Hemoglobin Trajectories",
    subtitle = paste0("Adjusted for mean baseline age; ITN use = ", itn_ref,
                      ". Points = observed group means.")
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

# Random Effects Distribution

```{r random-effects-dist}
# Histograms of child-level random intercepts and slopes 
p_ri <- ggplot(data.frame(intercept = re[, 1]), aes(x = intercept)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#3498DB", alpha = 0.6) +
  geom_density(linewidth = 0.8) +
  labs(x = "Random Intercept", y = "Density",
       title = "(A) Random Intercepts") +
  theme_minimal(base_size = 11)

p_rs <- ggplot(data.frame(slope = re[, 2]), aes(x = slope)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#E67E22", alpha = 0.6) +
  geom_density(linewidth = 0.8) +
  labs(x = "Random Slope (Month)", y = "Density",
       title = "(B) Random Slopes") +
  theme_minimal(base_size = 11)

p_ri + p_rs
```
